On The Impact of Atoms of Confusion in JavaScript Code

### 2021-04-25 4_Result.tex
additional review in the survey / MSR results

In this section we present the results of the 
interviews with the practitioners who agreed to participate
in the study. A total of 15 practitioners
took part of the interviews.
We collected information regarding programming
experience, familiarity with JavaScript, and their opinion
about the nine atom candidates---by
(a) presenting them pairs of code snippets whose behavior were identical
and then (b) asking them to compare the code snippets readability.

@@

\subsubsection*{The Participants' perceptions of the atom candidates}
Supporting the results of the survey, Table~\ref{tab:interview-results1}
 shows that for eight out of the
nine scenarios surveyed, the majority of the respondents agree that the version of the
code without the atom of confusion is easier to understand, and in no case the \emph{neutral}
ratio was higher than the option for the version without atoms of confusion.
%% An example entry is contained in Appendix~\ref{}. 
%% For a full listing of the code snippets, visit the paper repository. 
Only for the Ternary Operator atom candidate the participants
prefer the version of the code with the atom of confusion,
instead of the more cleaner version without the atom of confusion.
Even in this case, some participants who opted for
the \emph{left-hand-side} (with the atom of confusion) version
still believed that the \emph{right-hand-side} (without the atom of
confusion) version was more readable (see Figure~\ref{code:ternary}). 
The following quotes were extracted from the transcripts with
three interviewees:

@@

\begin{mq}
\emph{``I think [the \lhs version] is easier to understand, but [the \rhs version] is what I would write"}
\end{mq}

The atom candidate Ternary Operator also opens up
the possibility for a derivative construct that JavaScript
allows which can be rather confusing, and that is the nested
ternaries construct, in which the right-hand side of the a
ternary operator can be another ternary construct.
While nested ternaries remove the number of lines that
would be necessary to construct using nested if-then-else statements,
they can become quite taxing to understand.
Nested ternaries are a choice of atoms to be analyzed in future work.

@@

\begin{mq}
  \emph{``This one is interesting, because I have written
  code that looks like the [\lhs version (with the atom of confusion)],
  and my colleagues complained that it was difficult to understand.
  So nowadays I prefer to write code using the [\rhs version]."}
\end{mq}

Two of the atom candidates had unanimous preference for the
versions that did not contain the atoms: Comma Operator and
Omitted Curly Braces. Regarding the first one, we could
often notice during the interviews that
the \emph{\lhs} (with the atom of confusion) caused significant
confusion among the participants. Two remarks about
the \emph{Comma Operator} atom of confusion are listed below:

\begin{mq}
\emph{``I just learned that [the \lhs version of this code] is possible. I did not even know it worked''}
\end{mq}

@@

This is an interesting perspective, because such small, yet tricky program constructs and idioms, might hinder novice developers in the task of maintaining code; and one cannot make any assumption about the programmers' experience of who is going to maintain a code in a long run.

\begin{table}[!htb]
    \centering
    \caption{Interviews Round 1 - Summary of participants' preferences for code snippets \textsc{with} and \textsc{without} atoms of confusion (\textsc{aoc}). Participants were also allowed to choose \textsc{neutral} when they thought both sides were equally readable.}
    \label{tab:interview-results1}
    \begin{tabular}{lrrr}\toprule
      & \multicolumn{3}{c}{\textsc{Preference (\%)}} \\
      \cmidrule(lr){2-4}
         \textsc{atom}           & \multicolumn{1}{c}{\textsc{with aoc}}
                                      &  \multicolumn{1}{c}{\textsc{without aoc}}
                                               & \multicolumn{1}{c}{\textsc{neutral}} \\ \midrule
         Arithmetic as Logic     & 0  & 93.33  & 6.67  \\
         Assignment as Value     & 20 & 60     & 20    \\
         Comma Operator          & 0  & 100    & 0     \\
         Ternary Operator        & 60 & 26.67  & 13.3  \\
         Implicit Predicate      & 20 & 73.33  & 6.67  \\
         Logic as Control Flow   & 20 & 60     & 20    \\
         Omitted Curly Braces    & 0  & 100    & 0     \\
         Post Increment          & 20 & 73.33  & 6.67  \\
         Pre-Increment           & 40 & 46.67  & 13.33 \\ \midrule
         \textsc{overall}        & 20 & 70.37  & 9.63  \\
         \bottomrule
    \end{tabular}
\end{table}

@@

Moreover, such abstractions might lead a developer to start to program
using the language without actually understanding some of its basic aspects.
For instance, four participants argued that the language is not typed, two  believed the language runs exclusively on browsers, and two other developers stated that the language is natively asynchronous, all of which are incorrect statements. For the multi-purpose language that is one of the most, if not the most used one by practitioners,
a lack of understanding of the languages core
features may be a problem worth investigating in future studies.

@@

We mined \minedprojects open source JavaScript repositories to understand how often atoms of confusion arise in real software. We discarded in this analysis the atom Implicit Predicate,
because writing a CodeQL query for mining this particular atom is everything but trival.
As such, in this study
we only mined the eight atoms listed in Table~\ref{tab:occurrences-summary}.

Similarly to previous studies~\cite{DBLP:conf/msr/GopsteinZFC18}~\cite{DBLP:journals/ese/MedeirosLAAKRG19}, which investigates the prevalence of atoms of confusion in open source \clang and \cpplang projects, we found that atoms of confusion frequently arise in JavaScript open source systems. For instance, the 5 most frequently found atoms occur in at least 80\% of the projects.
Considering the extremes, atom Ternary Operator was found in all repositories,
while atom Comma Operator occurred in only 15\% of them (as seen in Table~\ref{tab:occurrences-summary}).

@@

% latex table generated in R 4.0.3 by xtable 1.8-4 package
% Tue Apr 13 09:34:35 2021
\begin{table}[ht]
\centering
\caption{Summary of atoms occurrences on our dataset}
\setlength\tabcolsep{2pt} % default value: 6pt
\label{tab:occurrences-summary}
\begin{tabular}{lrcc}
  \hline
Atom & Projects & Occurrences/KLOC & Category \\ 
  \hline
Implicit Predicate & 100\% & 19.89 &  highly used \\ 
  Ternary Operator & 100\% & 10.16 &  highly used \\ 
  Omitted Curly Braces & 91.67\% & 6.61 &  highly used \\ 
  Post Increment & 90.28\% & 5.62 &  highly used \\ 
  Pre Increment & 83.33\% & 1.04 & commonly used \\ 
  Assignment as Value & 81.94\% & 1.02 & commonly used \\ 
  Logic as Control Flow & 56.94\% & 0.95 & commonly used \\ 
  Automatic Semicolon Insertion & 33.33\% & 0.13 & commonly used \\ 
  Arithmetic as Logic & 23.61\% & 0.02 & little used \\ 
  Comma Operator & 15.28\% & 0.03 & little used \\ 
   \hline
\end{tabular}
\end{table}

Considering all JavaScript projects in our dataset, we found 
a total of {\color{red}\num{364873}} atoms of confusion, though three atoms are responsible for 87.97\% of this total: Ternary Operator, Omitted Curly Braces and the Post Increment. The remaining atoms
account for \num{24558} (12.03\%) of the total number of occurrences, and the Comma Operator and Arithmetic as Logic are the atoms that arise less frequently (0.001\% in total with 232 and 165 occurrences, respectively).

Regarding frequency, we proceeded with a classification of \emph{atom occurrence}
according to the thresholds of a previous
study~\cite{DBLP:journals/ese/MedeirosLAAKRG19}. As such, we found three
atoms as highly used (Ternary Operator, Omitted Curly Braces, and Post Increment),
three as commonly used (Pre-Increment, Assignment as Value, and Logic as Control Flow)
and two atoms as little used (Arithmetic as Logic and Comma Operator).
We also calculated the frequency of each atom per KLOC and found the atom
Ternary Operator as the most occurring in practice (10.16 occurrences per KLOC).
%In our dataset of JavaScript projects, 
%We found the Ternary Operator atom occurring more frequently than the Omitted Curly Braces, differently from what has been reported in a previous work~\cite{DBLP:conf/msr/GopsteinZFC18}. %

The results of our survey and interviews suggest that the Ternary Operator does not contribute significantly as a source of misunderstanding (increasing the number of wrong answers in 9\% of the cases, according to our survey). Nonetheless, the Post Increment Expression and Omitted Curly Braces atoms are listed in the top three sources of misunderstanding 
(see Table~\ref{tab:difference-correctness}). As such, fixing the the Post Increment and Omitted Curly Braces might solve most of the unclear code caused by atoms of confusion
in JavaScript projects.

>>>>

In this section we present the results of the 
interviews with the practitioners who agreed to participate
in the study. A total of 15 practitioners
took part of the interviews.
We collected information regarding programming
experience, familiarity with JavaScript, and their opinion
about the \na atom candidates---by
(a) presenting them pairs of code snippets whose behavior were identical
and then (b) asking them to compare the code snippets readability.

@@

\subsubsection*{The Participants' perceptions of the atom candidates}
Supporting the results of the survey, Table~\ref{tab:interview-results1}
 shows that for eight out of the
nine scenarios surveyed, the majority of the respondents agree that the version of the
code without the atom of confusion is easier to understand, and in no case the \emph{neutral}
ratio was higher than the option for the version without atoms of confusion.
%% An example entry is contained in Appendix~\ref{}. 
%% For a full listing of the code snippets, visit the paper repository. 
Only for the Ternary Operator atom candidate the participants
prefer the version of the code with the atom of confusion,
instead of the more cleaner version without the atom of confusion.
We show the code snippets for the Ternary Operator in Figure~\ref{code:ternary}.
Even for this atom candidate, some participants who preffered 
the \emph{left-hand-side} version (with the atom of confusion) 
still believed that the \emph{right-hand-side} version (without the atom of
confusion) was more readable.  
The following quotes were extracted from the transcripts with
three interviewees:

@@

%% \begin{mq}
%% \emph{``I think [the \lhs version] is easier to understand, but [the \rhs version] is what I would write"}
%% \end{mq}

%% The atom candidate Ternary Operator also opens up
%% the possibility for a derivative construct that JavaScript
%% allows which can be rather confusing, and that is the nested
%% ternaries construct, in which the right-hand side of the a
%% ternary operator can be another ternary construct.
%% While nested ternaries remove the number of lines that
%% would be necessary to construct using nested if-then-else statements,
%% they can become quite taxing to understand.
%% Nested ternaries are a choice of atoms to be analyzed in future work.

@@

\begin{mq}
  \emph{``This one is interesting, because I have written
  code that looks like the [\lhs version (with the atom of confusion)],
  and my colleagues complained that it was difficult to understand.
  Nowadays I prefer to write code using the [\rhs version]."}
\end{mq}

Two of the atom candidates had unanimous preference for the
versions that did not contain the atoms: Comma Operator and
Omitted Curly Braces. Regarding the first one, we could
often notice during the interviews that
the \emph{\lhs} (with the atom of confusion) caused significant
confusion among the participants. One remark about
the \emph{Comma Operator} atom of confusion is listed below:

%% \begin{mq}
%% \emph{``I just learned that [the \lhs version of this code] is possible. I did not even know it worked''}
%% \end{mq}

@@

This is an interesting perspective, because such a perhaps simple decision,  might hinder novice developers in the task of maintaining code; and one cannot make any assumption about the programmers' experience of who is going to maintain a code in a long run.

\begin{table}[!htb]
    \centering
    \caption{Summary of participants' preferences for code snippets \textsc{with} and \textsc{without} atoms of confusion (\textsc{aoc}). Participants were also allowed to choose \textsc{neutral} when they thought both sides were equally readable.}
    \label{tab:interview-results1}
    \begin{tabular}{lrrr}\toprule
      & \multicolumn{3}{c}{\textsc{Preference (\%)}} \\
      \cmidrule(lr){2-4}
         \textsc{atom}           & \multicolumn{1}{c}{\textsc{with aoc}}
                                      &  \multicolumn{1}{c}{\textsc{without aoc}}
                                               & \multicolumn{1}{c}{\textsc{neutral}} \\ \midrule
         Arithmetic as Logic     & 0  & 93.33  & 6.67  \\
         Assignment as Value     & 20 & 60     & 20    \\
         Comma Operator          & 0  & 100    & 0     \\
         Ternary Operator        & 60 & 26.67  & 13.3  \\
         Implicit Predicate      & 20 & 73.33  & 6.67  \\
         Logic as Control Flow   & 20 & 60     & 20    \\
         Omitted Curly Braces    & 0  & 100    & 0     \\
         Post Increment          & 20 & 73.33  & 6.67  \\
         Pre-Increment           & 40 & 46.67  & 13.33 \\ \midrule
         \textsc{overall}        & 20 & 70.37  & 9.63  \\
         \bottomrule
    \end{tabular}
\end{table}

@@

Moreover, such abstractions might lead a developer to start to program
using the language without actually understanding some of its basic aspects.
For instance, four participants argued that the language is not typed, two  believed the language runs exclusively on browsers, and two other developers stated that the language is natively asynchronous, all of which are incorrect statements.
%% For the multi-purpose language that is one of the most, if not the most used one by practitioners,
%% a lack of understanding of the languages core
%% features may be a problem worth investigating in future studies.

@@

We mined \minedprojects open source JavaScript repositories to understand how
often atoms of confusion arise in real software. 
Similarly to previous studies~\cite{DBLP:conf/msr/GopsteinZFC18,DBLP:journals/ese/MedeirosLAAKRG19}, which investigte
the prevalence of atoms of confusion in open source \clang and \cpplang projects, we found that atoms of confusion
frequently arise in JavaScript open source systems. For instance, the six most frequently found atoms
occur in at least 80\% of the projects. Considering the extremes, atom Ternary Operator was found in all repositories we mined,
while atom Comma Operator occurred in only 15\% of them (as seen in Table~\ref{tab:occurrences-summary}).

@@

% latex table generated in R 4.0.3 by xtable 1.8-4 package
% Tue Apr 13 09:34:35 2021
\begin{table}[ht]
\centering
\caption{Summary of atoms occurrences on our dataset}
\setlength\tabcolsep{2pt} % default value: 6pt
\label{tab:occurrences-summary}
\begin{tabular}{lrcc}
  \hline
Atom & Projects & Occurr./KLOC & Category \\ 
  \hline
Implicit Predicate & 100\% & 19.89 &  highly used \\ 
  Ternary Operator & 100\% & 10.16 &  highly used \\ 
  Omitted Curly Braces & 91.67\% & 6.61 &  highly used \\ 
  Post Increment & 90.28\% & 5.62 &  highly used \\ 
  Pre Increment & 83.33\% & 1.04 & commonly used \\ 
  Assignment as Value & 81.94\% & 1.02 & commonly used \\ 
  Logic as Control Flow & 56.94\% & 0.95 & commonly used \\ 
  Automatic Semicolon Insertion & 33.33\% & 0.13 & commonly used \\ 
  Arithmetic as Logic & 23.61\% & 0.02 & little used \\ 
  Comma Operator & 15.28\% & 0.03 & little used \\ 
   \hline
\end{tabular}
\end{table}

Considering all JavaScript projects in our dataset, we found 
a total of {\color{red}\num{364873}} atoms of confusion, though three atoms are responsible for 87.97\% of this total: Ternary Operator, Omitted Curly Braces, and the Post Increment. The remaining atoms
account for \num{24558} (12.03\%) of the total number of occurrences, and the Comma Operator and Arithmetic as Logic are the atoms that arise less frequently (0.001\% in total with 232 and 165 occurrences, respectively).

Regarding frequency, we proceeded with a classification of \emph{atom occurrence}
according to the thresholds of a previous
study~\cite{DBLP:journals/ese/MedeirosLAAKRG19}. As such, we found three
atoms as highly used (Ternary Operator, Omitted Curly Braces, and Post Increment),
four as commonly used (Pre-Increment, Assignment as Value,
Logic as Control Flow, and Automatic Semicolon Insertion),
and two atoms as little used (Arithmetic as Logic and Comma Operator).
We also calculated the frequency of each atom per KLOC and found the atom
Ternary Operator as the most occurring in practice (10.16 occurrences per KLOC).
%In our dataset of JavaScript projects, 
%We found the Ternary Operator atom occurring more frequently than the Omitted Curly Braces, differently from what has been reported in a previous work~\cite{DBLP:conf/msr/GopsteinZFC18}. %

The results of our survey and interviews suggest that the
Ternary Operator does not contribute significantly as a source of
misunderstanding (increasing the number of wrong answers in 9\% of the cases,
according to our survey). Nonetheless, the Post-Increment Expression
and Omitted Curly Braces atoms are listed in the top four sources of misunderstanding 
(see Table~\ref{tab:difference-correctness}). As such, fixing the
the Post-Increment and Omitted Curly Braces might solve most
of the unclear code caused by atoms of confusion in
JavaScript projects.
